Generating Random Logic Programs Using Constraint Programming

### 2020-01-02 paper/paper.tex
A better approach to symmetry reduction for clause heads

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}

@@

\def\multiset#1#2{\ensuremath{\left(\kern-.3em\left(\genfrac{}{}{0pt}{}{#1}{#2}\right)\kern-.3em\right)}}

@@

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$. All constraint variables in the model are integer
variables, but, e.g., if the integer $i$ refers to a logical variable $X$, we
will use $i$ and $X$ interchangeably.

@@

\begin{itemize}
\item Heads are too restrictive: $\variable{P}(X, Y, X)$ should be allowed.
\item A constraint for logical equivalence.
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

@@

Our definition is slightly more involved because we want to impose some
additional constraints, namely, that the variables in the atom are consecutive

\begin{definition} % mention: variables are always ordered, constants are not.
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item $\variable{predicate} \in \predicates \cup \{ \Box \}$, where $\Box$
    denotes a disabled clause.
  \item $\arrayd{variables}{|\variables{}|}{[0, \maxArity{}]}$: how many times
    each variable appears in the head atom.
  \item $\arrayd{constants}{\maxArity{}}{\constants{} \cup \{ \Box \}}$, where
    $\Box$ denotes that the position is reserved for a variable.
  \end{itemize}
  We define the \variable{predicate}'s \variable{arity} using the
  \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      0 & \text{if } \variable{predicate} = \Box\\
      \text{the arity of } \variable{predicate} & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  For each $v \in \variable{variables}$, $v \le \variable{arity}$.
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{constants}[i] = \Box.
  \]
\end{constraint}

\begin{constraint}
  Connecting the two lists:
  \[
    \maxArity{} - \variable{arity} + \sum_{v \in \variable{variables}} v
    = \begin{cases}
      \maxArity{} & \text{if } \variable{predicate} = \Box \\
      \variable{count}(\Box, \variable{constants}) & \text{otherwise}.
      \end{cases}
  \]
\end{constraint}

In \variable{variables}, all zeros must go after all non-zeros. For example,
if we have to pick one variable out of two, we must pick the first one.
\begin{constraint}
  For $i = 0, \dots, |\variables{}| - 2$, and $j = i + 1, \dots,
  |\variables{}| - 1$,
  \[
    \variable{variables}[i] \ne 0 \quad \text{or} \quad \variable{variables}[j]
    = 0.
  \]
\end{constraint}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (0, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1, -1) {$\neg$};
    \node (P) at (-1, -2) {$\variable{P}(X)$};
    \node[draw,circle,gray,text=black] (and) at (1, -1) {$\land$};
    \node (Q) at (0, -2) {$\variable{Q}(X)$};
    \node (R) at (2, -2) {$\variable{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \caption{A tree representation of the formula from \cref{example:formula}}
  \label{fig:example_tree}
\end{figure}

\begin{example} \label{example:formula}
  Let $\maxNumNodes{} = 8$. Then $\neg\variable{P}(X) \lor (\variable{Q}(X)
  \land \variable{P}(X))$ corresponds to the tree in \cref{fig:example_tree} and
  can be encoded as:
  \begin{alignat*}{8}
    \variable{treeStructure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
    \variable{treeValues} &= [{\lor}, &&{\neg}, &&{\land}, \variable{P}(&&X), \variable{Q}(&&X), \variable{P}(&&X), &&\top, &&\top],\\
    \variable{numNodes} &= 6, && && && && && && &&\\
    \variable{numTrees} &= 3. && && && && && && &&
  \end{alignat*}
  In the rest of this section, we will describe how the elements of
  $\variable{treeValues}$ are encoded and list a series of constraints that make
  this representation unique.
\end{example}

@@

\begin{example}
  Let $\maxArity{} = 2$, $\predicates{} = [\variable{P}, \dots]$, $\arities{}
  = [1, \dots]$, and $X \in \variables{}$. Then the node representing atom
  $\variable{P}(X)$ has:
  \begin{align*}
    \variable{name} &= \variable{P},\\
    \variable{arguments} &= [X, 0],\\
    \variable{arity} &= 1.
  \end{align*}
\end{example}

@@

\begin{constraint}
  Each predicate gets at least one clause. Let $P = \{ h.\variable{predicate}
  \mid h \in \variable{headsOfClauses} \}$. Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      \variable{numPredicates} & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.

@@

\section{Counting Programs}

@@

\section{The Independence Constraint}

% TODO: define independence of predicates
% Formal definition (here and in the predicate invention paper): two
% predicates are independent if all of their groundings are independent.

>>>>

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}

@@

%\relpenalty=10000
%\binoppenalty=10000
\def\multiset#1#2{\ensuremath{\left(\kern-.3em\left(\genfrac{}{}{0pt}{}{#1}{#2}\right)\kern-.3em\right)}}

@@

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$. All constraint variables in the model are integer
variables, but, e.g., if the integer $i$ refers to a logical variable $X$, we
will use $i$ and $X$ interchangeably. All indices start at zero.

@@

\begin{itemize}
\item Heads are too restrictive: $\mathsf{P}(X, Y, X)$ should be allowed.
\item A constraint for logical equivalence. An algorithm to reduce each tree to
  some kind of normal form.
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

@@

Our definition is slightly more involved because we want to eliminate some
symmetries.

\begin{definition}
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item a $\variable{predicate} \in \predicates \cup \{ \Box \}$, where $\Box$
    denotes a disabled clause,
  \item and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}}$.
  \end{itemize}
\end{definition}

\begin{definition}
  The \variable{predicate}'s $\variable{arity} \in [0, \maxArity{}]$ can then be
  defined using the \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      0 & \text{if } \variable{predicate} = \Box\\
      \text{the arity of } \variable{predicate} & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{arguments}[i] = 0.
  \]
\end{constraint}

\begin{definition} We define an array \variable{occurrences} of $|\variables{}|$
  sets such that, for all $v \in \variables{}$, $\variable{occurrences}[v]$ is a
  subset of $\{ 0, \dots, \maxArity{} - 1\}$ that stores the indices of
  \variable{arguments} at which $v$ is located. It is defined by the
  \variable{setsIntsChanneling} constraint equivalent to the following
  condition: for $i = 0, \dots, \maxArity{} - 1$ and $v \in \variables{}$,
  \[
    \variable{arguments}[i] = v \quad \iff \quad i \in \variable{occurrences}[v].
  \]
\end{definition}

\begin{definition}
  We define a $|\variables{}| \times 2$ matrix $\mathbf{M}$ with domain $[0,
  \maxArity{}]$ such that for $v = 0, \dots, |\variables{}| - 1$,
  \begin{align*}
    \mathbf{M}_{v,0} &= |\variable{occurrences}[v]|.\\
    \mathbf{M}_{v,1} &= \begin{cases}
      \min \variable{occurrences}[v] & \text{if } \variable{occurrences}[v] \ne \emptyset\\
      0\footnotemark & \text{otherwise.}
    \end{cases}
  \end{align*}
\end{definition}
\footnotetext{This could be any fixed value in the domain.}

\begin{constraint}
  We can then impose the \variable{lexChainLessEq} constraint on $\mathbf{M}$
  which means that for $v = 1, \dots, |\variables{}| - 1$,
  \[
    |\variable{occurrences}[v - 1]| \le |\variable{occurrences}[v]|,
  \]
  and if $|\variable{occurrences}[v - 1]| = |\variable{occurrences}[v]| > 0$,
  then
  \[
    \min \variable{occurrences}[v - 1] \le \min \variable{occurrences}[v].
  \]
\end{constraint}

\begin{example}
  Let $\maxArity{} = 4$, $\mathsf{P} \in \predicates{}$, $\variables{} = \{
  X, Y, Z \}$. Then $\mathsf{P}(Z, Y, Z)$ would be represented as:
  \begin{align*}
    \variable{predicate} &= \mathsf{P},\\
    \variable{arguments} &= [Z, Y, Z, 0],\\
    \variable{arity} &= 3,\\
    \variable{occurrences} &= [\emptyset, \{ 1 \}, \{ 0, 2 \}],\\
    \mathbf{M} &= \begin{bmatrix}
      0 & 0\\
      1 & 1\\
      2 & 0
    \end{bmatrix}.
  \end{align*}
\end{example}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (0, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1, -1) {$\neg$};
    \node (P) at (-1, -2) {$\mathsf{P}(X)$};
    \node[draw,circle,gray,text=black] (and) at (1, -1) {$\land$};
    \node (Q) at (0, -2) {$\mathsf{Q}(X)$};
    \node (R) at (2, -2) {$\mathsf{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \caption{A tree representation of the formula from \cref{example:formula}}
  \label{fig:example_tree}
\end{figure}

\begin{example} \label{example:formula}
  Let $\maxNumNodes{} = 8$. Then $\neg\mathsf{P}(X) \lor (\mathsf{Q}(X)
  \land \mathsf{P}(X))$ corresponds to the tree in \cref{fig:example_tree} and
  can be encoded as:
  \begin{alignat*}{8}
    \variable{treeStructure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
    \variable{treeValues} &= [{\lor}, &&{\neg}, &&{\land}, \mathsf{P}(&&X), \mathsf{Q}(&&X), \mathsf{P}(&&X), &&\top, &&\top],\\
    \variable{numNodes} &= 6, && && && && && && &&\\
    \variable{numTrees} &= 3. && && && && && && &&
  \end{alignat*}
  In the rest of this section, we will describe how the elements of
  $\variable{treeValues}$ are encoded and list a series of constraints that make
  this representation unique.
\end{example}

@@

\begin{example}
  Let $\maxArity{} = 2$, $\predicates{} = [\mathsf{P}, \dots]$, $\arities{}
  = [1, \dots]$, and $X \in \variables{}$. Then the node representing atom
  $\mathsf{P}(X)$ has:
  \begin{align*}
    \variable{name} &= \mathsf{P},\\
    \variable{arguments} &= [X, 0],\\
    \variable{arity} &= 1.
  \end{align*}
\end{example}

@@

\begin{constraint}
  Each predicate gets at least one clause. Let
  \[
    P = \{ h.\variable{predicate} \mid h \in \variable{headsOfClauses} \}.
  \]
  Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      \variable{numPredicates} & \text{otherwise.}
    \end{cases}
  \]
  Here, $\variable{nValues}(P)$ counts the number of unique values in $P$.
\end{constraint}

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body. % TODO: can this be turned into lexChainLessEq?

@@

\section{Counting Programs}
% TODO: update to account for the new definition of head

@@

\section{Predicate Independence}

\begin{definition}
  Let $\mathscr{P}$ be a probabilistic logic program. Its \emph{predicate
    dependency graph} is a directed graph $G_{\mathscr{P}} = (V, E)$ with the
  set of nodes $V$ consisting of all predicates in $\mathscr{P}$. We add an edge
  from predicate $\mathsf{P}$ to predicate $\mathsf{Q}$ if there is a clause in
  $\mathscr{P}$ with $\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the
  body.
\end{definition}

\begin{definition}
  Let $\mathsf{P}$ be a predicate in a program $\mathscr{P}$. The
  \emph{dependencies} of $\mathsf{P}$ is the smallest set $D_{\mathsf{P}}$ such
  that:
  \begin{itemize}
  \item $\mathsf{P} \in D_{\mathsf{P}}$,
  \item for every $\mathsf{Q} \in D_{\mathsf{P}}$, the nodes with arrows to
    $\mathsf{Q}$ in $G_{\mathscr{P}}$ are all in $D_{\mathsf{P}}$.
  \end{itemize}
\end{definition}

\begin{definition}
  Two predicates $\mathsf{P}$ and $\mathsf{Q}$ are \emph{independent} if
  $D_{\mathsf{P}} \cap D_{\mathsf{Q}} = \emptyset$.
\end{definition}
